---
title: "ERA5CleanData"
author: "Anna Talucci"
date: "2024-02-13"
output: html_document
---

##Mathes edits 
```{r clear environment, include=FALSE}
rm(list=ls())
```

# Overview

Take extracted ERA 5 data from Google Earth Engine calculate DOY to estimate (est) ALT from.

ERA5 data was extract by year.

# Packages

```{r}
library(tidyverse)
library(lubridate)
library(sf)
library(data.table)
```



# Data

## vector of data files
```{r}
f <- list.files(path = "/Users/kmathes/Desktop/PermafrostThaw/TransientThaw/Output/ERA5_dailyTemperature_EE_expanded/",
                pattern = "*.csv", full.names = TRUE)
```

```{r}
f
```

## Individual files
```{r}
f2001 <- read.csv(f[1], header = TRUE) 
f2005 <- read.csv(f[2], header = TRUE) 
f2006 <- read.csv(f[3], header = TRUE) 
f2007 <- read.csv(f[4], header = TRUE) 
f2008 <- read.csv(f[5], header = TRUE) 
f2009 <- read.csv(f[6], header = TRUE) 
f2010 <- read.csv(f[7], header = TRUE) 
f2011 <- read.csv(f[8], header = TRUE) 
f2012 <- read.csv(f[9], header = TRUE) 
f2013 <- read.csv(f[10], header = TRUE) 
f2014 <- read.csv(f[11], header = TRUE) 
f2015 <- read.csv(f[12], header = TRUE) 
f2016 <- read.csv(f[13], header = TRUE) 
f2017 <- read.csv(f[14], header = TRUE) 
f2018 <- read.csv(f[15], header = TRUE) 
f2019 <- read.csv(f[16], header = TRUE) 
f2020 <- read.csv(f[17], header = TRUE) 
```

## View DF
```{r}
f2001
```
```{r}
names(f2001)
```

## Process
1. pivot longer so column of temperature and date
2. Group by site_id
3. Select dates after day of thaw measurement
4. Select only rows with temp above 0 (zero)
5.  apply calculation

"resBiomNm", "resBiomeNm", "resEcId", "resId", "resName", "resRelm", "wwfBiom", "wwfEcId", "wwfId", "wwfName", "wwfNum", "wwfRelm"  






# Functions to clean 
```{r}
selectColDf = function(x)  {
  x %>%
    dplyr::select(plotId, siteId, cntryId, lastNm, submtNm, year, month, day, biome, distur, fireYr, paired, gtProbe, hitRock, srfH2O, topoPos, slope, vegCvr, msrDoy, msrType, msrDpth, resBiom, resName,  prmExtn, prdDpth, predDoy,  contains("temperature_2m")) %>%
    mutate(id = row_number()) -> x
  return(x)
}
```

```{r}
cleanDf = function(x)  {
  x %>%
  pivot_longer(
    cols = starts_with("X"),
    names_to = "DateTemp",
    values_to = "TempC",
    values_drop_na = TRUE) %>% 
  separate(DateTemp, c("tempDate", NA)) %>% 
    mutate(plotId = as.character(plotId)) %>%
    mutate(across(tempDate, ~ . |> str_extract("\\d+") |> as.character())) %>%
    mutate(tempDate=ymd(tempDate)) %>%
    mutate(TempDoy = yday(tempDate)) -> x
  return(x)
}
```


##  Apply functions to to each year

### Select Column Function

```{r}
df2001 = f2001 %>% selectColDf()
df2005 = f2005 %>% selectColDf()
df2006 = f2006 %>% selectColDf()
df2007 = f2007 %>% selectColDf()
df2008 = f2008 %>% selectColDf()
df2009 = f2009 %>% selectColDf()
```

```{r}
df2010 = f2010 %>% selectColDf()
df2011 = f2011 %>% selectColDf()
df2012 = f2012 %>% selectColDf()
df2013 = f2013 %>% selectColDf()
df2014 = f2014 %>% selectColDf()
```

```{r}
df2015 = f2015 %>% selectColDf()
df2016 = f2016 %>% selectColDf()
df2017 = f2017 %>% selectColDf()
df2018 = f2018 %>% selectColDf()
df2019 = f2019 %>% selectColDf()
```

```{r}
df2020 = f2020 %>% selectColDf()

```


### Clean function

Note: the warning is referencing parts of the name being discarded in the separate process. This can be ignored for these chunks of code.

```{r}
( clean2001 = df2001 %>% cleanDf() )
```

```{r}
clean2005 = df2005 %>% cleanDf()
clean2006 = df2006 %>% cleanDf()
clean2007 = df2007 %>% cleanDf()
clean2008 = df2008 %>% cleanDf()
clean2009 = df2009 %>% cleanDf()
```

```{r}
clean2010 = df2010 %>% cleanDf()
clean2011 = df2011 %>% cleanDf()
clean2012 = df2012 %>% cleanDf()
clean2013 = df2013 %>% cleanDf()
clean2014 = df2014 %>% cleanDf()
```

```{r}
clean2015 = df2015 %>% cleanDf()
clean2016 = df2016 %>% cleanDf()
clean2017 = df2017 %>% cleanDf()
clean2018 = df2018 %>% cleanDf()
clean2019 = df2019 %>% cleanDf()
```

```{r}
clean2020 = df2020 %>% cleanDf()

```

# Bind clean 

Combine yearly data frames; rename columns that were truncated by EE. 

```{r}
(combineAll = dplyr::bind_rows(clean2001, clean2005, clean2006, clean2007, clean2008, clean2009, clean2010, clean2011, clean2012, clean2013, clean2014, clean2015, clean2016, clean2017, clean2018, clean2019, clean2020)) 


```


```{r}
names(combineAll)
```

```{r}
unique(combineAll$lastNm)
```


# write to csv

```{r}
write.csv(combineAll,file = "/Users/kmathes/Desktop/PermafrostThaw/TransientThaw/Output/ERA5_dailyTemperature_EE_expanded/ERA_dailyTemperature_expanded_summary.csv", row.names = FALSE)
```



